---
title: "1. Introduction to R"
author: "Adnan Shroufi"
date: "2025-02-05"
output: html_document
---

```{r setup, echo=TRUE, eval=TRUE}
if (!requireNamespace("rprojroot", quietly = TRUE)) {install.packages("rprojroot")}
knitr::opts_knit$set(root.dir = find_root(criterion = is_rstudio_project))
```

# Introduction

Welcome to the Dig Data Step-up Challenge!

In this session we will:

- learn the basics of R Studio
- learn some simple Base R functions
- become more familiarised with the dataset in the challenge

## Why Use R?

Firstly, you might be asking, why bother using R in the first place, compared with a tool like Microsoft Excel.
While Excel is a great tool with lots of benefits, it does have some limitations. These include:

- Excel can be slow when working with large files
- Going over the same steps in Excel every day/week/month to generate the same report/analysis is time consuming and repetitive.

Using R can can overcome these issues for the following reasons:

- R can work with large files far more easily and quickly
- A single R script can do all of the time consuming and repetitive tasks for you

## Markdown Documents

All of the R sections of the Step-up Challenge are written in a 'markdown' format.
This is kind of like a 'notebook' in Python.

Basically, you have text, interspersed with cells containing code, whcih can be run.
You can run a cell by clicking on the green arrow to the right hand side of a cell.

The below code will print some text into the bottom-left pane of R Studio.

```{r}
print("Welcome to the R Step-up Challenge!")
```

## R Studio Overview

R Studio is known as an Integrated Development Environment (IDE).
Within this IDE you can load data, write code, see the output from code, and inspect charts, and so on.
The R Studio IDE is made up on 4 panes, or sections.
Each one has a purpose and can be expanded or re-sized as you would like.

Top left: Source (where you write code)
Bottom left: Console (where code is evaulated by R)
Top right: Environment (to see and inspect data and objects)
Bottom right: Navigation (to navigate and view files and plots) 

## Directories and Loading Data

So, R is a programming language with lots of benefits.
However, if you're using it for the first time, there will be a lot of new things to learn.

With programming, you have to tell a computer exctaly what to do.
And if there is a slight mistake or mis-spelling with an instrauction, it basically won't run.

When loading data, you need to first know what directory you are in.
Think of your computer like a giant digital filing cabinet with lots of drawers and folders inside.

In R Studio, getwd() is like asking, "Which drawer and folder am I currently working in?"
We can do this now.

```{r}
getwd()
```

We can now list all of the files that are inside the folder or location we are working from (the directory).

```{r}
list.files()
```

Next, we want to load some data from the data folder.
Using the navigation pane (bottom-right) you can see the Data folder.

Inside that is ssome data called 'STEP_UP_REGIONAL_ANTIDEPRESSANTS.Rds'
An 'Rds' is an R-specific data type, that is more compressed than a CSV file.
It also allows formatting to saved within the object, that a CSV file would be unable to do. 

We will load this with the 'readRDS' function.
If we want to read this data and save it as an object within our environment pane (top right), we have to *assign* the object a name.
We must spell the file correctly (all code and objects are case-sensitive in R) and assign it a name (in this instance just 'data').

```{r}
data = readRDS("STEP_UP_REGIONAL_ANTIDEPRESSANTS.Rds")
```

So, the above does not work, because we haven't specified the folder to load the data from.
We need to load the data form the 'Data' folder.
And we need to use a forward slash ('/') between the folder location and the object name.

```{r}
data = readRDS("Data/STEP_UP_REGIONAL_ANTIDEPRESSANTS.Rds")
```

## Global Environment

We can now view our data in the Global Environment.
You can't view an Rds object within the file explorer, like you can with a CSV file.
We just need to click on the object name.

Inside the object view we can:

- Sort columns numerically or alphabetically
- Filter columns by a value or piece of text (string) 

The name of the type of object in global environment is a *dataframe*.
A *dataframe* is like a table you might see in Excel or Google Sheets. 
It’s a way of organizing data in rows and columns.
This is where each column holds a specific type of information, and each row holds data for a different item or record.

## What is the actual data we are using?

This data is from the NSHBSA Open Data Portal, and is called the 'Prescription Cost analysis' dataset.

https://opendata.nhsbsa.net/dataset/prescription-cost-analysis-pca-monthly-data

This data is open source, which means:

- We can share it openly
- It has no security concerns

This is a monthly dataset that describes how many medicines were prescribed across all GP Practices in england, per NHS Region, and what they cost.
This is 'real data', so actually describes real anti-depressant prescribing in England.
We have merged, simplified and filtered all these monthly files into a single dataset.

This dataset now contains 6 columns:

*YM*: The year and month, in the format YYYYMM, where 202401 is the same as January 2024. There are 46 year-month values in the data.
*YEAR*: The year in the format YYYY. There are 4 years-worth of prescribing information in the dataset.
*REGION*: The NHS Region. There are 7 regions in the data.
*DRUG*: The name of the anti-depressant medicine. There are 32 of these in the data.
*ITEMS*: How many items were prescribed.
*COST*: The combined cost for all those items. 

In a sentence we could describe this dataset as:

- Per English NHS Region and per year-month, the volume and cost of each antidepressant drug prescribed.

And what we are going to do with this data is:

- Understand national and regional prescribing volumes and costs
- Understand national and regional prescribing trends
- Understand monthly and annual trends
- And finally, maybe even predict future monthly anti-depressant prescribing volumes

Some of the following base R functions will help us have an overview of the data content and structure. 

## Some basic base R Functions

We will now look at some 'Base R' functions.
'Base R' are the in-built functions within R.
There are lots of them, but we will just look at a few.

The str() function lets you know the 'structure' of an object.
You just need to 'wrap' the function around the *data* object.

```{r}
str(data)
```

We can see how many rows and columns there are.
We can also see what type of data is contained in each column.
These are 'Factor', 'chr', 'int' and 'num'.

Each is described below:

*Factor*: A factor is a type of data in R that represents categories or groups. 
Imagine you're sorting people into different groups based on their favorite color (like red, blue, and green). 
Each person belongs to one of these groups, and the groups themselves are factors. 
For example, "red" is a category or level of the "color" factor.

*Character*: This refers to text or words. 
When you have things like names or descriptions (like "John" or "pizza"), those are character data. 
It's like when you're writing a report or document of some kind. — you're using characters (letters and words).

*Integer*: An integer is just a whole number. 
So, it's any number that doesn’t have a decimal, like 1, 5, or 100. 
If you think about counting objects (like apples or books), you'd use integers because you're counting whole things, not parts.

*Numeric*: Numeric data can be any kind of number, including those with decimals. 
For example, 2, 3.5, and 10.75 are all numeric. 
It’s like when you're measuring things, like the height of someone (5.6 feet) or the temperature (23.7°C).

## Specifying Columns Within a Dataframe

If we type the name of an R object and then *$* sign, the possible column names will then be accessible.
That way we can specify a column name.
We often need to feed a column name using a *$* sign into a function.

If we feed the numeric and integer columns into summary(), we can better undertsand the distribution of the data.

```{r}
summary(data$ITEMS)
summary(data$COST)
```

We can feed the character or factor columns into the table() function to get a count of records per character or factor entry.
The number under each character or factor describes how many records each one has. 

```{r}
table(data$REGION)
table(data$DRUG)
```

A final few Base functions.
First is head(), this displays the first few rows of a dataframe, if you want to quickly check the content.

```{r}
head(data)
```

Finally, is:

- how to check the number of columns in a dataframe
- how to check the number of rows in a dataframe
- how to check the number of levels in a factor type column

```{r}
ncol(data)
nrow(data)
nlevels(data$YM)
nlevels(data$YEAR)
```

## Summary

So, we know how to find out which directory we are in and how to load in some data.

We have an initial understanding of what this data is actually about and it's contents. 

And up until now we have used the following R functions:

- print()
- getwd()
- list.files()
- readRDS()
- str()
- summary()
- table()
- head()
- ncol()
- nrow()
- nlevels()

In the next section we will look at R libraries, and in particular the *dplyr* library.
