---
title: "Visualising Data"
author: "Adnan Shroufi"
date: "2025-02-11"
output: html_document
---

```{r setup, echo=TRUE, eval=TRUE}
if (!requireNamespace("rprojroot", quietly = TRUE)) {install.packages("rprojroot")}
knitr::opts_knit$set(root.dir = rprojroot::find_root(criterion = rprojroot::is_rstudio_project))
```

# Introduction

Visualising is an important tool in both analytics and data science.
Visualising data can help us better understand data and see trends in data, amongst other things.
We will use an R package called *highcharter* to do this.
We will first install this package using *install.packages*.

```{r}
install.packages("highcharter")
```

Thee highcharter is free to use for students, and some public sector organisations.
Private companies have to pay an annual subscription fee to use the package, which is eomthig to be aware of. 
We will then install this and thee dplyr packages into this session using the *library* function.

```{r}
library(highcharter)
library(dplyr)
```

And load in the same dataset using *readRDS*.

```{r}
data = readRDS("Data/STEP_UP_REGIONAL_ANTIDEPRESSANTS.Rds")
```

## Overview

The highcharter package is an interesting and useful package because:

- You can easily create some very nice looking charts
- You can also create (more complicated) production-level charts 

Within highcharter, we will only be using the *hchart* function, to easily create some very ince looking charts.
We can create a huge number of charts with hchart just using 3 things:

- Dataset
- Chart Type
- Chart Aesthetic (or how variables in your data are mapped to the chart visual properties).

Here is an exmaple of the kind of charts you can create using hchart and very little code.

```{r}
browseURL("https://jkunst.com/highcharter/articles/hchart.html")
```

## Chart Types

In this session we will look at 3 kinds of chart types:

- Vertical bar charts
- Horizontal bar charts
- Line charts

The aesthetic for each of these chart types is just defining what information you want displayed on the chart (numerical) y-axis and x-axis.

## Chart Elements

When visualising data in R (or any other programming language), there are usually a few steps you would through.

1. Aggregate the data
2. Create a basic plot
3. Refine or develop the plot

Point 1. is very important.
If you want to visualise data in a certain way, you must first aggregate the data in a certain way.
The best way to demonstrate this is through an example.

# EXample 1: Aggregate Data

Let's say I want to create a bar chart that shows the cost of Fluoxetine hydrochloride prescribing nationally each year.
Because there are only 4 years, I would need a dataframe with only 4 rows to do this, with an ITEM value per YEAR. 
To generate this data, I would have to do the following aggregation:

1. Filter the DRUG to Fluoxetine hydrochloride
2. Group by YEAR
3. Summarise and sum by ITEMS
4. Ungroup

The code would look like this:

```{r}
df_bar = data %>% 
  filter(DRUG == "Fluoxetine hydrochloride") %>% 
  group_by(YEAR) %>% 
  summarise(COST = sum(COST)) %>% 
  ungroup()
```

## Example 1: Basic Chart

Using this data, I can then define 2 of the 3 things I need to create a basic hchart:

1. The data (df_bar)
2. The chart type ("column")
2. The Chart aesthetic (YEAR vs COST)

I want to use the df_bar object, and I want to create a bar chart where the x-axis is the region name, and the y-axis is the total number of items.
The following code would do this.

```{r}
hchart(df_bar, "column", hcaes(YEAR, COST))
```

# Example 1: Refine Chart

This is a start, although the chart needs some work, the following steps would make it look nicer.

1. Round the data to whole number of pounds
2. Add some axis labels
3. Add a title

We will do these refinements one by one.
We can also 'chain' or 'pipe' highcharter commands together suing the '%>%' operator.

First, we can either round the data as part of the aggregation, or within the chart code, using the *round* function.
For simplicity we can do this within the chart function, although generally speaking it is good to do this within the data aggregation stage. 

```{r}
hchart(df_bar, "column", hcaes(YEAR, round(COST)))
```

Second, we can add some x-axis and y-axis labels using the *hc_xAxis* and *hc_yAxis* functions.
Each of these functions have multiple parameters.
And some of these parameters even have multiple parameters within themselves.
For example, the *title* parameter has multiple parameters, which must in turn be defined within a list.
Our x-axis and y-axis code will not get much more complicated than this, so don't worry about this code syntax.

```{r}
hchart(df_bar, "column", hcaes(YEAR, round(COST))) %>% 
  hc_xAxis(title = list(text = "Year")) %>% 
  hc_yAxis(title = list(text = "Millions of Pounds (£)"))
```

Finally, we need to give the chart an appropriate title.
The title should describe exactly what the showing is showing.

```{r}
hchart(df_bar, "column", hcaes(YEAR, round(COST))) %>% 
  hc_xAxis(title = list(text = "Year")) %>% 
  hc_yAxis(title = list(text = "Millions of Pounds (£)")) %>% 
  hc_title(text = "The annual cost of national anti-depressant prescribing from 2021 until 2024")
```
  
## Example 2: Aggregate the Data

Let's say I want to create a bar chart that shows how much Setraline Hydrochloride was prescribed in each region in 2024.
Because there are only 7 regions, I would need a dataframe with only 7 rows to do this, with an ITEM value per REGION. 
To generate this data, I would have to do the following aggregation:

1. Filter the DRUG to Setraline Hydrochloride
2. Filter the YEAR to 2024
3. Group by REGION
4. Summarise and sum by ITEMS
5. Ungroup

The code would look like this:

```{r}
df_bar_two = data %>% 
  filter(DRUG == "Sertraline hydrochloride", YEAR == 2024) %>% 
  group_by(REGION) %>% 
  summarise(ITEMS = sum(ITEMS)) %>% 
  ungroup() %>% 
  arrange(ITEMS)
```

## Example 2: Basic Chart

I want to create a horizontal bar chart.
This is because some region names are very long.
In highcharter:

Vertical bar chart = "column"
Horizonatal bar chart = "bar

Once again, I need 3 things to create my basic chart:

1. The data (df_bar_two)
2. Chart type ("bar")
3. The Chart aesthetic (REGION vs ITEMS)

Now, even through visually, the x-axis will be ITEMS, highcharter expects the numerical value of a "bar" or "column" chart to be the y-axis.
With this in mind, I want to use the df_bar_two object, and I want to create a bar chart where the x-axis is the region name, and the y-axis is the total number of items.
The following code would do this.

```{r}
hchart(df_bar_two, "bar", hcaes(REGION, ITEMS)) 
```

## Example 2 : Refine Chart

The values don't need rounding, so this chart just needs:

1. Add axis labels
2. Add a title

These will both be done together is one go.
Remember, the y-axis label is for the numerical axis with a hchart type "column" or "bar".

```{r}
hchart(df_bar_two, "bar", hcaes(REGION, ITEMS)) %>% 
  hc_xAxis(title = list(text = "Region")) %>% 
  hc_yAxis(title = list(text = "Number of sertraline hydrochloride items")) %>% 
  hc_title(text = "The total regional sertraline hydrochloride prescribing volume in 2024")
```

We can use the exact same code, except just change chart type to "column".
As can be seen, we can switch between chart types with little effort (if our data is aggregated and/or formatted appropriately).

```{r}
hchart(df_bar_two, "column", hcaes(REGION, ITEMS)) %>% 
  hc_xAxis(title = list(text = "Region")) %>% 
  hc_yAxis(title = list(text = "Number of sertraline hydrochloride items")) %>% 
  hc_title(text = "The total regional sertraline hydrochloride prescribing volume in 2024")
```

## Example 3: Aggregate Data

Next we will create a line chart, that shows the monthly cost of all anti-depressant prescribing in London, for the entire time period.
This is going to be a single value per month, so the aggregated data should have the same number of rows as the number of months in the data. 
To generate this data, I would have to do the following aggregation:

1. Filter the REGION to London
2. Group by YM
4. Summarise and sum by COST
5. Ungroup

The code would look like this:

```{r}
df_line = data %>% 
  filter(REGION == "LONDON") %>% 
  group_by(YM) %>% 
  summarise(COST = sum(COST)) %>% 
  ungroup()
```

## Example 3: Basic Chart

Once again, I need 3 things to create my basic chart:

1. The data (df_line)
2. The chart type ("line")
3. The Chart aesthetic (YM vs COST)

I want to use the df_line object, and I want to create a line chart where the x-axis is the year-month, and the y-axis is the prescribing cost.
The following code would do this.

```{r}
hchart(df_line, "line", hcaes(YM, COST))
```

## Example 3: Refine Chart

This is a start, although the chart, once again, needs some work, the following steps would make it more a more useful chart.

1. Adjust the minimum y-axis value.
2. Add axis labels
3. Add a title

We will first set the y-axis minimum value to zero.
This is optional, and depends on the scale of the chart.
In this instance, the y-axis minimum being zero may give you a better sense of the changing prescribing volumes. 

```{r}
hchart(df_line, "line", hcaes(YM, COST)) %>% 
  hc_yAxis(min = 0)
```

We can now add the axis labels and a title.
The y-axis information can be inserted into the same hc_yAxis function that has set the y-axis minimum value.
Alternatively, the y-axis title information can be specified within a seperate function call.
Do whichever you find easier.
The 2 code chunks below both create the same chart.

```{r}
# Identical chart No. 1
hchart(df_line, "line", hcaes(YM, COST)) %>% 
  hc_yAxis(min = 0) %>% 
  hc_yAxis(title = list(text = "Number of antidepressants")) %>% 
  hc_xAxis(title = list(text = "Year-month")) %>% 
  hc_title(text = "The monthly number of antidepressants prescribed in London from January 2021 until December 2024")

# Identical chart No. 2
hchart(df_line, "line", hcaes(YM, COST)) %>% 
  hc_yAxis(min = 0, title = list(text = "Number of antidepressants")) %>% 
  hc_xAxis(title = list(text = "Year-month")) %>% 
  hc_title(text = "The monthly number of antidepressants prescribed in London from January 2021 until December 2024")
```

# Example 4: Aggregate the Chart

In this chart, we will compare the national monthly prescribing volume of 2 drugs.
To do this, we need to use a new logical operator, the *|* logical operator, which means *or*.

The code below we filter the data for either of the 2 drugs mentioned

```{r}
df_line_two = data %>% 
  filter(DRUG == "Sertraline hydrochloride" | DRUG == "Amitriptyline hydrochloride")
```

We can do a quick summarise to check what DRUG remain in the dataset.
As we can see, the filter has only retained the 2 specified drugs.
If you wanted, you could use multiple *|* operators, for example to retain only 3 or 4 DRUG. 

```{r}
df_line_two %>% 
  group_by(DRUG) %>% 
  summarise(ITEMS = sum(ITEMS)) %>% 
  ungroup()
```

So, we want to filter the data the data for only these 2 DRUG, but then aggregate COST for each YM.
To do this we need to:

1. Filter either DRUG (using OR)
2. Group_by YM and DRUG (because we want values for each DRUG)
3. Sum COST (and round to nearest pound)

```{r}
df_line_two = data %>% 
  filter(DRUG == "Sertraline hydrochloride" | DRUG == "Amitriptyline hydrochloride") %>% 
  group_by(YM, DRUG) %>% 
  summarise(COST = round(sum(COST))) %>% 
  ungroup()
```
## Example 4: Basic chart

We now want to create a hchart, but with one line for each DRUG.
To do this, we need to use the *group* parameter, where *group = DRUG*.
In effect, this is saying create one line for each DRUG.

In highcharter, each set of data points is called a *series*.
If you click on a series name, underneath the chart, you can remove/add a series. 

```{r}
hchart(df_line_two, "line", hcaes(YM, COST, group = DRUG))
```

## Example 4: Refine Chart

We will refine in a few ways:

1. Change y-axis minimum value 
2. Add axis labels
3. Add a title
4. Create a shated tooltip

We will do the first 3 steps in one go.

```{r}
hchart(df_line_two, "line", hcaes(YM, COST, group = DRUG)) %>% 
  hc_yAxis(min = 0, title = list(text = "Prescribing cost in millions (£)")) %>% 
  hc_xAxis(title = list(text = "Year-month")) %>% 
  hc_title(text = "The monthly cost of national sertraline hydrochloride and amitriptyline hydrochloride prescribing from January 2021 until December 2024")
```

If the mouse cursor hovers over each line, seperate values can be viewed within the tooltip (one per series).
We can view both series together using the following code *hc_tooltip(shared = TRUE)*. 

```{r}
hchart(df_line_two, "line", hcaes(YM, COST, group = DRUG)) %>% 
  hc_yAxis(min = 0, title = list(text = "Prescribing cost in millions (£)")) %>% 
  hc_xAxis(title = list(text = "Year-month")) %>% 
  hc_title(text = "The monthly cost of national sertraline hydrochloride and amitriptyline hydrochloride prescribing from January 2021 until December 2024") %>% 
  hc_tooltip(shared = TRUE)
```

## 'Piping' straight into hchart and different chart types

We need to aggregate data before using it within hchart.
However, you can actually 'pipe' the aggregation outout straight into hchart if you want.

We will create a vertical bar chart that shows the annual cost of all antidepressant prescribing to demonstrate this.
First we will aggrgeate the data.

```{r}
df_bar_three = data %>% 
  group_by(YEAR) %>% 
  summarise(COST = round(sum(COST))) %>% 
  ungroup()
```

We can then visualise the data.

```{r}
hchart(df_bar_three, "column", hcaes(YEAR, COST))
```

Alternatively, we can 'pipe' the aggregation straight into hchart.
If you do this, leave the data parameter blank, as chart knows the piped output is the visualisation input.

```{r}
data %>% 
  group_by(YEAR) %>% 
  summarise(COST = round(sum(COST))) %>% 
  ungroup() %>% 
  hchart("column", hcaes(YEAR, COST))
```

Finally, we will see how many different charts we can make with a single dataframe.
This highlights the flexibility and power of hchart.
Note, some of charts could be made more effective, if the data were aggregated more appropriately for that singular chart type..

```{r}
df_multichart = data %>% 
  group_by(YEAR, REGION) %>% 
  summarise(
    COST = round(sum(COST)),
    ITEMS = sum(ITEMS)
    ) %>% 
  ungroup()
```

Here are multiple charts using the above dataframe (although one has an additional filter and then piped into hchart).

```{r}
# Vertical bar chart grouped by region
hchart(df_multichart, "column", hcaes(YEAR, COST, group = REGION))

# Horizontal bar chart grouped by region
hchart(df_multichart, "bar", hcaes(YEAR, COST, group = REGION))

# Horizontal bar chart grouped by region
hchart(df_multichart, "area", hcaes(YEAR, COST, group = REGION)) %>% 
  hc_plotOptions(area = list(stacking = "normal"))

# Annual line chart grouped by region
hchart(df_multichart, "line", hcaes(YEAR, COST, group = REGION))

# Same as above but with a curvy line 
hchart(df_multichart, "spline", hcaes(YEAR, COST, group = REGION))

# Heatmap of cost per region per year, rounded to nearest 1000 (see negative value within round function)
hchart(df_multichart, "heatmap", hcaes(YEAR, REGION, value = round(COST, -3))) %>% 
  hc_plotOptions(heatmap = list(dataLabels = list(enabled = TRUE)))

# scatter chart of regional cost vs items in 2024 (with each point being a region)
df_multichart %>% 
  filter(YEAR == 2024) %>% 
  hchart("scatter", hcaes(COST, ITEMS, group = REGION))
```

The above charts are just to show you the range of things you can do with very little additional tweaking and/or additional code.
Again, here are the examples again of what you can do with a basic hchart function.
You can complete all exercises and reports just using "bar", "column" and "line" chart types, so don't worry about the above charts you are seeing for the first time.

```{r}
browseURL("https://jkunst.com/highcharter/articles/hchart.html")
```

## Summary

Visualising data is a key daya analytics and data science skill.
It starts with aggregating data, in a way that it is appropriately formatted for the type of chart you want.
It's then a case of firstly creating a base plot.
After that, it's about refining the plot appearance and format.

These are all the highcharter-related functions we have used in this session.

- hchart()
- hc_yAxis()
- hc_xAxis
- hc_title()
- hc_tooltip()
- hc_plotOptions (additional code for new ccharts at the end)
- round()

## Exercise

Now try the exercise within the Visualising Data folder.
This is found within the 'visualising_data_exercise.r' script.
This is a 'regular' R script rather than a markdown document.
Write your code to answer each question within the script.
