---
title: "Data Metrics and Insight"
author: "Adnan Shroufi"
date: "2025-02-12"
output: html_document
---

```{r setup, echo=TRUE, eval=TRUE}
if (!requireNamespace("rprojroot", quietly = TRUE)) {install.packages("rprojroot")}
knitr::opts_knit$set(root.dir = rprojroot::find_root(criterion = rprojroot::is_rstudio_project))
```

# Introduction

In this section we will start generating prescribing-related metrics.
So far we have just reported on data, and well done if you have made it this far.
Because regions are different sizes with different quantities of prescribing, it is hard to compare their prescribing meaningfully.
Metric generation will open up this new area of insight.
We will first load our key packages.

```{r}
library(dplyr)
library(highcharter)
```

## Data

In the extension work, which is this session plus the next exercise, we will use a new dataset.
This is an expanded version of the original data we were using.

```{r}
data = readRDS("Data/EXTENSION_STEP_UP_REGIONAL_ANTIDEPRESSANTS.Rds")
```

The following filter and select 'reduces' this data down the size of original dataset.
(We can also use -select to de-select a column).

```{r}
original_data = data %>% 
  filter(BNF_SECTION == "Antidepressant drugs") %>% 
  select(-BNF_CHAPTER, -BNF_SECTION)
```

So we can see that the original data differs in a few ways:

- It only contains antidepressant drugs
- It doesn't conatin BNF Chapter and Section information

The BNF stands for the *British National Formulary*.
The BNF is structured hierarchically into Chapters, Sections and Chemical Substances (Drugs).
(There are also Paragraphs and Sub-Paragraphs but we won't cover those).

For example:

- Amitriptyline hydrochloride is an actual antidepressant *DRUG*
- It is one of many *DRUG* within the 'Antidepressant drugs' *BNF_SECTION*
- It is one of many *BNF_SECTION* within the '04: Central Nervous System' *BNF_CHAPTER*
- And there are 23 *BNF_CHAPTER* (although very little prescribing stems from some of the chapters)

## Exploratory Data Analysis (EDA)

EDA is a technique where you explore a dataset, in order to better understand it.
Data visualisation is an effective way to do basic EDA.
Using the skills and techniques we can create a few charts to do this high-level exploration.

First, we can see the total prescribing per chapter within a single year.
We can see that the Central Nervous System (CNS) chapter had the 2nd most amount of prescribing, which contains the antidepressant section.
We can also see that some BNF chapters have (comparatively) very little prescribing.

```{r}
data %>% 
  filter(YEAR == 2023) %>% 
  group_by(BNF_CHAPTER) %>% 
  summarise(ITEMS = sum(ITEMS)) %>% 
  ungroup() %>% 
  hchart("bar", hcaes(BNF_CHAPTER, ITEMS))
```

We can also see what are top 20 BNF sections our of all the chapters.
We will also filter the data to just one year. 
It can be seen the antidepressants section had the 2nd most amount of prescribing (from all sections from all chapters).

```{r}
data %>% 
  filter(YEAR == 2023) %>% 
  group_by(BNF_SECTION) %>% 
  summarise(ITEMS = sum(ITEMS)) %>% 
  ungroup() %>% 
  top_n(20, ITEMS) %>% 
  arrange(ITEMS) %>% 
  hchart("bar", hcaes(BNF_SECTION, ITEMS))
```

Finally, for context, we can use a line chart to see the monthly volume of national prescribing (for all chapters).
We do not need to filter by year as we want all months included.
We will set the y-axis minimum value to zero to get a better sense of scale.

```{r}
data %>% 
  group_by(YM) %>% 
  summarise(ITEMS = sum(ITEMS)) %>% 
  ungroup() %>% 
  hchart("line", hcaes(YM, ITEMS)) %>% 
  hc_yAxis(min = 0)
```

## Metric Generation

In order to go beyond tallying numbers and recording data, we need to think how to meaningfully compare data.
For example, the below chart shows the antidepressant section prescribing volume for each region across the whole timeframe.

```{r}
data %>% 
  filter(BNF_SECTION == "Antidepressant drugs") %>% 
  group_by(YM, REGION) %>% 
  summarise(ITEMS = sum(ITEMS)) %>% 
  ungroup() %>% 
  hchart("line", hcaes(YM, ITEMS, group = REGION)) %>% 
  hc_yAxis(min = 0)
```


Now, someone might say from this data:

'The North East prescribes the most antidepressants and London the least.'

Which would be factually true, but this doesn't take a lot of things into account, including:

- The total population of each region
- The population of each region by age band (as those aged 65+ receive more prescribing)
- The total prescribing volume in each region
- Other factors, such as the deprivation profile of each region

To overcome this, we can create a metric that *standardise* these values.
This means the the metrics have a common scale enabling to be compared meaningfully.
The following metric would do this:

- What *percentage* of prescribing per month in each region is for antidepressants

To calculate this, we would need to do the following:

- Calculate the monthly antidepressant volume per region
- Calculate the total prescribing volume per region
- Divide these numbers and turn into a percentage

## Creating new Features (mutate)

To create new features (i.e. a column or variable) within a datatset we need to use *mutate*.
Mutate is one of the most R functions you will come across.
This either creates a new variables, or 'overwrites' an old variable.

For example, the below is creating a new variable called COST2, which is the COST column multiplied by 2.
Only a handful of columns are pre-selected before doing this mutate.

```{r}
data %>% 
  select(YM, REGION, DRUG, ITEMS, COST) %>% 
  mutate(COST2 = COST * 2)
```

The below code is doing the same, but this giving not changing the column name.
In this instance, the multiplied value will replace the initial value.
We use the * symbol to multiply values or columns.

```{r}
data %>% 
  select(YM, REGION, DRUG, ITEMS, COST) %>% 
  mutate(COST = COST * 2)
```

With mutate, it is therefore important to think about if you want to:

- create a new column
- overwrite the content of an existing column

## Conditional logic

When creating new columns using mutate, it is often useful to corporate conditional logic.
The simplest way to do this is using *ifelse*.

An ifelse works in the following way:

1. Specify a condition
2. If this is true then do one thing
3. If this is false then do something else

We will also introduce 2 new logical operators.
Previously we saw

- equals to *==*
- not equals to *!=*

We also have (for numerical columns):

- more than *>*
- more than or equal to *>=*
- less than *<*
- less than *<=*

We will use these, along with an ifelse, to create a new column called VOLUME.
This is purely to explain the ifelse operation, this is not a recognised threshold or category type.

The below code does the following:

1. Check if the (regional, monthly and drug-level) item count is less than 1000
2. If it less than 1000 create the label "LOW", and populate the VOLUME column with this value 
3. If it equal or more than 100 create the label "HIGH", and populate the VOLUME column with this value 

```{r}
data %>% 
  select(YM, REGION, DRUG, ITEMS) %>% 
  mutate(VOLUME = ifelse(ITEMS < 1000, "LOW", "HIGH")) 
```

The output to an ifelse is only ever 2 things, A or B.
If you need conditional logic with 2 or more conditions, you can use something (in R) called a *case_when*.
We will not cover this, but feel free to look it up if you are curious.
You can check the R documentation that accompanies a function by doing the following:

```{r}
?case_when
```

We could use the same conditional logic, but have the resulting column formatted differently.
For example, the below:

1. Check if the (regional, monthly and drug-level) item count is less than 1000
2. If it less than 1000, generate a value of 0, and populate the ITEMS_NEW column with this value 
3. If it equal or more than 100, take whatever value is currently in the ITEMS column, and populate the ITEMS_NEW column with this value 

If you are using an ifelse and mutate together, first think about what you want your resulting column to look like.
Then, think about what condition, or conditional logic, would be able to create such an output.
The ifelse in this instance is kind of being used as a 'rounding function'.

```{r}
data %>% 
  select(YM, REGION, DRUG, ITEMS) %>% 
  mutate(ITEMS_NEW = ifelse(ITEMS < 1000, 0, ITEMS)) 
```

This code creates a new column that splits the data into 2 groups, prescribing from LONDON, and everything else.

```{r}
data %>% 
  select(YM, REGION, DRUG, ITEMS) %>% 
  mutate(REGION_LONDON = ifelse(REGION == "LONDON", "LONDON", "ALL OTHER REGIONS")) 
```

The below code creates a new column AD_ITEMS. If items are in the 'Antidepressant drugs' BNF_SECTION, then we take the existing item count. Otherwise we generate a zero value. 

```{r}
data %>% 
  select(YM, REGION, BNF_SECTION, DRUG, ITEMS) %>% 
  mutate(AD_ITEMS = ifelse(BNF_SECTION == "Antidepressant drugs", ITEMS, 0))
```

## Metric Generation using Conditional Logic

The above ifelse may not seen that useful, but we can actually use to to calculate the earlier specified proposed metric, which was:

- What *percentage* of prescribing per month in each region is for antidepressants

We can use the above code then do the following:

- Group by region and year-month
- Add up all the ITEMS
- Add up all the AD_ITEMS

We will assign the output from this into a variable so we can have a look at it in the global environment.

```{r}
ad_pct = data %>% 
  mutate(AD_ITEMS = ifelse(BNF_SECTION == "Antidepressant drugs", ITEMS, 0)) %>% 
  group_by(YM, REGION) %>% 
  summarise(
    AD_ITEMS = sum(AD_ITEMS),
    ITEMS = sum(ITEMS)
  ) %>% 
  ungroup()
```

We then use mutate again to divide these columns, to create a percentage (as a decimal).
We can call this new column whatever we want, but will go for AD_PCT (antidepressant percent).
The */* operator is used to divide, and we need to divide AD_ITEMS by total ITEMS.

```{r}
ad_pct = data %>% 
  mutate(AD_ITEMS = ifelse(BNF_SECTION == "Antidepressant drugs", ITEMS, 0)) %>% 
  group_by(YM, REGION) %>% 
  summarise(
    AD_ITEMS = sum(AD_ITEMS),
    ITEMS = sum(ITEMS)
  ) %>% 
  ungroup() %>% 
  mutate(AD_PCT = AD_ITEMS / ITEMS)
```

We can change this decimal to a value out of 100 by multiplying it by 100.
We can then round this to 2 d.p. (decimal point), using the *round* function.

```{r}
ad_pct = data %>% 
  mutate(AD_ITEMS = ifelse(BNF_SECTION == "Antidepressant drugs", ITEMS, 0)) %>% 
  group_by(YM, REGION) %>% 
  summarise(
    AD_ITEMS = sum(AD_ITEMS),
    ITEMS = sum(ITEMS)
  ) %>% 
  ungroup() %>% 
  mutate(AD_PCT = round(100 * AD_ITEMS / ITEMS, 2)) 
```

Finally, we can visualise this monthly percentage value by region.
Because we have multiple groups, we nee to specify *group = REGION*.
In this instance, we might want to *not* set the y-axis minimum value to zero, so that the regional trend lines can be better differentiated.

So, from this standardised data we can see that out of all prescribing, London routinely had the lowest percentage of monthly antidepressant prescribing (out of all prescribing).
There could be a huge number of reasons why this is the case (such as other medicines being in more demand) so we wouldn't be able to jump any conclusions here.
Nevertheless, this is an interesting chart, with the London trend line being notably different to others.


```{r}
hchart(ad_pct, "line", hcaes(YM, AD_PCT, group = REGION)) %>% 
  hc_xAxis(title = list(text = "Year-month")) %>% 
  hc_yAxis(title = list(text = "Percentage of antidepressant prescribing (%)")) %>% 
  hc_title(text = "The monthly percentage of antidepressant prescribing, out of all prescribing, per region from 202301 until 202412")
```

Finally, we will create a diferent using the same conditional logic, although a diferent group_by.
This will look at the percentage of antidepressant prescribing per region, just in 2023 (so ending up with 1 value per region).
To do this, we need to remove YM from the group_by.
We will assign this to a new object so we can inspect it.


```{r}
ad_pct_year = data %>% 
  filter(YEAR == "2024") %>% 
  mutate(AD_ITEMS = ifelse(BNF_SECTION == "Antidepressant drugs", ITEMS, 0)) %>% 
  group_by(REGION) %>% 
  summarise(
    AD_ITEMS = sum(AD_ITEMS),
    ITEMS = sum(ITEMS)
  ) %>% 
  ungroup()
```

We can then round as before, plus order the calculated percentage value (so the bars in our final chart are ordered).

```{r}
ad_pct_year = data %>% 
  filter(YEAR == "2024") %>% 
  mutate(AD_ITEMS = ifelse(BNF_SECTION == "Antidepressant drugs", ITEMS, 0)) %>% 
  group_by(REGION) %>% 
  summarise(
    AD_ITEMS = sum(AD_ITEMS),
    ITEMS = sum(ITEMS)
  ) %>% 
  ungroup() %>% 
  mutate(AD_PCT = round(100 * AD_ITEMS / ITEMS, 2)) %>% 
  arrange(AD_PCT)
```
  
We can then create a simple plot (without axis labels or a title) to show this information.

```{r}
hchart(ad_pct_year, "bar", hcaes(REGION, AD_PCT))
```

We can finally add some axis labels and a title.
Remember, when using a hchart "bar" style chart, the y-axis is the horizontal numerical value. 
When giving a chart a title, try to give one that describes exactly is being shown.

```{r}
hchart(ad_pct_year, "bar", hcaes(REGION, AD_PCT)) %>% 
  hc_xAxis(title = list(text = "Region")) %>% 
  hc_yAxis(title = list(text = "Percentage of antidepressant prescribing (%)")) %>% 
  hc_title(text = "The percentage of antidepressant prescribing, out of all prescribing, per region in 2023")
```

## Statistical Inference - Numerical

We can generate some basic summary statistics, about the distribution of values within a dataset, using the following functions:

- min() = minimum mvalue
- max() = maximum value
- median() = middle value within an ordered dataset
- mean() = the 'average' value (all values added together and divided by number of values)
- quantile(0.25) = lower quartile (Also known as the 25th percentile, the value below which 25% of the data falls)
- quartile(0.75) = upper quartile (Also known as the 75th percentile, the value below which 75% of the data falls)

You  may recognise these terms from boxplots (which we will look at later).

We can use these within summarise, either with or without a group_by.
If we wanted regional-level summary statistics, we would have to group_by region before calculating these statistics.
These resulting values would be in relation to the monthly percentage of antidepressant prescribing, for each region, from 202101 until 202412.
When calculating such statistics it is good to remind yourself what these values are actually referring to!

```{r}
ad_pct %>% 
  group_by(REGION) %>%
  summarise(
    min = min(AD_PCT),
    Q1 = quantile(AD_PCT, 0.25),
    median = median(AD_PCT),
    Q3 = quantile(AD_PCT, 0.75),
    max = max(AD_PCT),
    mean = mean(AD_PCT)
  ) %>% 
  ungroup()
```

The ad_pct_year data only has 7 values, if we wanted to calculate these values *across* regions, we would change the data and remove the group_by.
In this instance, the resulting values would be in relation to the regional percentage of antidepressant prescribing in 2023.
Again, when calculating such statistics it is good to remind yourself what these values actually mean.

```{r}
ad_pct_year %>% 
  summarise(
    min = min(AD_PCT),
    Q1 = quantile(AD_PCT, 0.25),
    median = median(AD_PCT),
    Q3 = quantile(AD_PCT, 0.75),
    max = max(AD_PCT),
    mean = mean(AD_PCT)
  )
```

## Statistical Inference - Visual

Finally, we can also use data visualisation as a means to generate some statistical inference.
We will use a boxplot, to calculate the following:

- Maximum value
- Upper quartile: Also known as the 75th percentile, the value below which 75% of the data falls. Represented by the upper boundary of the "box" in a boxplot
- Median: Also known as the 50th percentile, the middle value when the data is sorted. The middle line inside the "box" in a boxplot represents the median
- Lower quartile: Also known as the 25th percentile, the value below which 25% of the data falls. Represents the lower boundary of the "box" in a boxplot.
- Minimum value

The easiest way to do is using a slightly different highcharter function.
This is *hcboxplot* rather than *hchart* (technically this is a deprecated function but it still works so we will use it).
We just need to specify the following:

- The numerical column we are interetsed
- Our grouping variaable (optional)

One caveat, is that highcharter automatically identifies outliers and does not include them in the boxplot whisker 'minimum'.
To calculate the actual minimum, use the numerical method above.

We will first create a boxplot for our *ad_pct_year* object.
There is no grouping variable here.
It is important to give this a suitable title, otherwise someone looking at it would have no idea what is going on.
We can add a y-axis label too (the y-axis is always the numerical axis in highcharter, even if it the horizontal axis).

This is how you would create a boxplot without a grouping variable, just feeding a single column of numerical values.
In practice, a simple table of values (i.e. the actual ad_pct_year object itself) might be more useful than this.

```{r, warning=FALSE}
hcboxplot(x = ad_pct_year$AD_PCT) %>% 
  hc_title(text = "Distribution of the annual percentage of antidepressant prescribing from each region in 2023") %>% 
   hc_yAxis(title = list(text = "Percentage of antidepressant prescribing (%)"))
```

The above boxplot becomes far more informative interpretable if we add a grouping variable.
Again, in practice, you might want to focus only on a calendar (or financial) years-worth of data, unless there was a reason to focus on 4 years-worth of monthly values.

```{r, warning = FALSE}
hcboxplot(x = ad_pct$AD_PCT, var = ad_pct$REGION) %>% 
  hc_title(text = "Distribution of the monthly percentage of antidepressant prescribing for each region, from 202101 until 202412") %>% 
  hc_yAxis(title = list(text = "Percentage of antidepressant prescribing (%)"))
```

## Summary

We haven't introduced many new function, but *mutate* and *ifelse* are both incredibly powerful and versatile.
You can complete a lot of useful tasks just using these, such as creating standardised prescribing metrics.

And we calculate a variety of summary statistics, either with or without a group_by, using *min*, *max*, *mean*, *median* and *quartile*.
And the *hcboxplot* functions is a really easy way to visualise these summary statistics.

All the new functions and logical operators from this sessions are listed below.

- mutate()
- ifelse()
- round()
- min()
- max()
- median()
- mean()
- quartile()
- hcboxplot()
- *<=*
- *>=*

## Exercise

Now try the exercise within the Data Metrics and Insight folder.
This is found within the 'data_metrics_and_insight_exercise.r' script.
This is a 'regular' R script rather than a markdown document.
Write your code to answer each question within the script.
