---
title: 'Longitudinal Analysis of Antidepressant Prescribing Volume and Cost'
author: "YOUR NAME HERE"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true 
    toc_depth: 3  
    number_sections: true 
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_knit$set(root.dir = rprojroot::find_root(criterion = rprojroot::is_rstudio_project))
```

```{r, include=FALSE}
library(dplyr)
library(highcharter)
library(tidyr)

data = readRDS("Data/STEP_UP_REGIONAL_ANTIDEPRESSANTS.Rds")
```


# Task

You are a Public Health Analyst and you have been asked to write a report.
This report will describe and analyse antidepressant prescribing, looking at volume and cost, both nationally and regionally.
The title of this report is 'Longitudinal Analysis of Antidepressant Prescribing'.
Longitudinal just means the analysis is over an extended period of time, which we have with 4 years of monthly data.
The report is to be split into two sections:

Part One: To set the context for the report, this will be overall national and regional figures 
Part Two: The will be followed-up by a more exploratory analysis that delves into antidepressant prescribing cost trends.
Part Two Extension: Those attempting the extension task can then look at creating standardised metrics to understand more specific prescribing patterns.


# Part One

Part One is a directed analysis and the same content needs to be covered by both streams.
These required content for Part One is: 

1. Create two vertical bar charts for comparison.
   First, create a bar chart showing the total annual antidepressant prescribing  (items).
   Second, create a bar chart that shows the total antidepressant prescribing cost.
   Describe the trend of each chart and compare the charts against each other.
   
2. Create two tables for comparison.
   First, create a table showing the total annual antidepressant prescribing per region (items, pivot_wider).
   Second, create a table showing the annual antidepressant prescribing cost per region (pivot_wider). 
   Describe some regional changes and contrasts between 2021 and 2024.
   
3. By now you may have noticed there are contrasts between antidepressant prescribing volumes and costs.
   Next, we will create two horizontal bar charts for comparison.
   First, create an ordered bar chart showing the 10 most prescribed antidepressants in the most recent year (items).
   Second, create an ordered bar chart that shows antidepressants with the greatest total prescribing cost in the most recent year.
   Highlight and describe any differences between the charts. 
   
Try and make text between sections flow, so the report reads well.
The findings from part one may inform ho you approach part two.


# Part Two

Now you have a good understanding around national longitudinal antidepressant prescribing volume and cost trends.
You will be given various topics or points to explore, and you will decide what approach or charts best does this.
To supplement the initial analysis, you will now:

- First, look at the monthly cost of antidepressant prescribing per BNF Paragraph, for the most recent year.
- Then, explore the data and find the antidepressant drugs that are driving the monthly cost trends.
- Then give a high-level summary to your work and findings.


# Part Two Extension

The extension is only to be attempted if you completed the data_metrics_and_insights learning material and exercises. 
The analyses within this section will delve a bit deeper into antidepressant prescribing costs.

- First calculate the average cost per item, per BNF paragraph (4), within the BNF antidepressant section, across the whole time frame.
- Then select one or two drugs with a notable monetary impact, and calculate what their monthly cost amounts to, out of all monthly prescribing costs.
- The above may have to be displayed as two charts.
- Finally, feel free to add any other analyses or charts you feel could be useful or appropriate! 
- Then give a high-level summary to your work and findings.


# 'Knit' Markdown

When you have written all your text and the code to generate all your charts and tables, the report can now be 'knit'.
This means it can be converted into a html or pdf file, which you can then export and use as you like.
Click on the 'knit' icon at the top of the screen.
If you click the arrow cursor, you can choose the format of the output.

If you do *not* want code included within the report, then specify as below:

```{r, include=FALSE}
data %>% 
  group_by(YEAR) %>% 
  summarise(ITEMS = sum(ITEMS)) %>% 
  ungroup()
```

Using '#', '##', an '###' can automatically structure your document into headings and sub-headings.
If you are unsure about this, just use single '#' to denote each new section.  

# Heading 1

ABC

## Heading 2

ABC

### Heading 3

ABC
